Competing Bandits: \\ The Perils of Exploration under Competition

### 2018-10-31 paper/competing_bandits.tex
somewhat major revision
updates to paper

\usepackage{tikz} % SW: Added for plots
\usepackage{subfiles}
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath,amssymb,amsfonts}
%\usepackage{natbib} % cannot use with AAAI! A.S.

@@

We ask whether better exploration algorithms are incentivized under competition. We run extensive numerical experiments in a stylized duopoly model in which two firms deploy multi-armed bandit algorithms and compete for myopic users.  We find that duopoly and monopoly tend to favor a primitive ``greedy algorithm" that does not explore, whereas a temporary monopoly (a duopoly with an early entrant) may incentivize better bandit algorithms. Our findings shed light on the ``first-mover advantage" in the digital economy.
\end{abstract}

@@

\subfile{content/Introduction}

@@

>>>>

\usepackage{tikz} % SW: Added for plots
\usepackage{subfiles}
\usepackage[toc,page]{appendix}
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath,amssymb,amsfonts}
%\usepackage{natbib} % cannot use with AAAI! A.S.

@@

We ask whether better exploration algorithms are incentivized under competition. We run extensive numerical experiments in a stylized duopoly model in which two firms deploy multi-armed bandit algorithms and compete for myopic users.  We find that duopoly and monopoly tend to favor a primitive ``greedy algorithm" that does not explore \gaedit{and leads to low consumer welfare}, whereas a temporary monopoly (a duopoly with an early entrant) may incentivize better bandit algorithms \gaedit{and lead to higher consumer welfare. Our findings shed light on the first-mover advantage in the digital economy by exploring the role that data can play as a barrier to entry in online markets. We show that a data advantage may serve as a significant barrier to entry when exploration has significant reputational consequences for entrants.}
\end{abstract}

@@

\subfile{content/introduction}

@@

\subfile{content/conclusion}

\subfile{content/appendix_for_one_version}

### 2018-10-31 paper/content/conclusion.tex
updates to paper

\gaedit{In this paper we considered a setting where firms simultaneously learn from and compete for users. We showed that competition may not always induce firms to commit to algorithms that engage in exploration and that this leads to welfare losses for consumers. The primary mechanism behind this was that exploration may have short-term reputational consequences that lead to naive algorithms winning in competition. Allowing one firm to have a head-start, or a first-mover advantage, leads to the first-mover having an incentive to play traditionally ``good" bandit algorithms and having significantly better welfare for consumers. Finally, we attempted to isolate the effect that the data advantage of an incumbent has on its ability to garner a considerable market share. Contrary to previous studies and discussions which do not consider competition dynamics or endogenizing data collection, we find that the data advantage is substantial in generating significant market power. Additionally we show that this effect is strongest when an incumbent performs exploration in the incumbency period and generates a high-quality dataset on actions that may be difficult for an entrant to generate due to the pressure of competition.}

\gacomment{Not sure what else to put in the conclusion?}

>>>>

\gaedit{In this paper we considered a setting where firms simultaneously learn from and compete for users. We showed that competition may not always induce firms to commit to algorithms that engage in exploration resulting in welfare losses for consumers. The primary mechanism behind this was that exploration may have short-term reputational consequences that lead to naive algorithms winning in competition that are bad in the long-run. Allowing one firm to have a head-start, or a first-mover advantage, leads to the first-mover having an incentive to play traditionally ``good" bandit algorithms and having significantly better welfare for consumers. Finally, we attempted to isolate the effect that the data advantage of an incumbent has on its ability to garner a considerable market share. Contrary to previous studies and discussions which do not consider competition dynamics or endogenous data collection, we find that the data advantage is substantial in generating significant market power. Additionally we show that this effect is strongest when an incumbent performs exploration in the incumbency period and generates a high-quality dataset on actions that may be difficult for an entrant to generate due to the pressure of competition.}

### 2018-10-31 paper/content/introduction.tex
remove this line
updates to paper

\documentclass[../competing_bandits.tex]{subfiles}
\begin{document}

@@

Platforms routinely deploy A/B tests, and are increasingly adopting  more sophisticated exploration methodologies based on \emph{multi-armed bandits}, a well-known framework for exploration and making decisions under uncertainty. While deploying ``better" learning algorithms for exploration would improve performance, this is not necessarily beneficial under competition, even putting aside the deployment/maintenance costs. In particular, excessive experimentation may hurt platform's reputation and decrease market share in the near term. This would leave the learning algorithms with less users to learn from, which may further degrade platform's performance relative to competitors who keep learning and improving from \emph{their} users, and so forth. \gaedit{This ``data feedback" loop has been discussed in policy circles and in the business literature, but, to our knowledge, our paper is the first to formally provide a model that allows us to understand this effect.}

@@

We consider a \emph{permanent duopoly} in which both firms start at the same time, as well as \emph{temporary monopoly}: a duopoly with an early entrant. Accordingly, the intensity of competition in the model varies from ``permanent monopoly" (just one firm) to ``\gadelete{early entry}\gaedit{incumbent}" (first-mover in duopoly) to permanent duopoly to ``\gadelete{late entry}\gaedit{entrant" (late arriver in duopoly)}. \gaedit{We find that, even in a duopoly, in equilibrium competition incentivizes firms to play a ``greedy algorithm" that does not explicitly explore and even more so if the firm is a late entrant in a market.} This algorithm also prevails under monopoly, simply because it tends to be easier to deploy. Whereas a temporary monopoly incentivizes more advanced exploration algorithms that perform better in the long run. \gaedit{We find that consumer welfare is highest under temporary monopoly and that even increasing the number of firms in the market past a duopoly makes consumers weakly worse off.}

@@

\gaedit{Our findings show that the data feedback loop arises either when a firm deploys a ``naive" greedy learning algorithm (under simultaneous entry) or has an initial temporary monopoly in the market. The findings on temporary monopoly shed light on the ``first-mover advantage" phenomenon in the digital economy. Being first in the market in the digital economy gives firms free data to learn from (a ``data advantage") as well as a more definite, and possibly better reputation compared to an entrant (a ``reputation advantage").

Even though, under our model, consumers are best off when there is a ``temporary" monopoly, a recent subject of debate in policy circles is whether the data advantage alone poses a sufficiently powerful barrier to entry to warrant anti-trust interest. While \cite{varian2018artificial} and \cite{bajari2018impact} argue that, from a statistical perspective, data has decreasing returns to scale and therefore does not serve as a significant competitive advantage, they focus on the problem in isolation. We argue that, to reason about this question, one needs to consider what the impact of better data has in terms of competition dynamics and when data collection is endogenous and has consequences, especially when firms need to learn while competing. Therefore, we run additional experiments that attempt to isolate the impact of the data advantage and reputation advantage. We find that the data advantage alone still leads to a significant incumbency advantage. Additionally, the data advantage tends to be stronger when the incumbent commits to a more advanced bandit algorithm. This highlights that, when thinking about data as a barrier to entry, data quality and not just quantity matters. In particular, the data advantage may be larger and more relevant when the incumbent possesses data that it gets via exploration of potentially sub-optimal actions that would be costly for the entrant to gather upon entry.}

@@

The study of whether data can give a significant competitive advantage has been discussed theoretically in \cite{varian2018artificial, lambrecht2015can} and empirically in \cite{bajari2018impact}. These studies consider the effect of additional data in terms of helping towards solving a machine learning problem in isolation and argue that it does not play a large role, whereas we ask how additional data helps while firms simultaneously solve a machine learning problem and competing for a market of consumers. In our setting, the competition dynamics from our model lead to different conclusions about the competitive advantage that a data advantage can give a firm. Additionally, the first-mover advantage has been well-studied in economics and marketing \cite{kerin1992first}. }

>>>>

\documentclass[../competing_bandits_with_appendix.tex]{subfiles}
\begin{document}

@@

Platforms routinely deploy A/B tests, and are increasingly adopting  more sophisticated exploration methodologies based on \emph{multi-armed bandits}, a well-known framework for exploration and making decisions under uncertainty. While deploying ``better" learning algorithms for exploration would improve performance, this is not necessarily beneficial under competition, even putting aside the deployment/maintenance costs. In particular, excessive experimentation may hurt platform's reputation and decrease market share in the near term. This would leave the learning algorithms with less users to learn from, which may further degrade platform's performance relative to competitors who keep learning and improving from \emph{their} users, and so forth.

@@

We consider a \emph{permanent duopoly} in which both firms start at the same time, as well as \emph{temporary monopoly}: a duopoly with a \gaedit{first-mover}. Accordingly, the intensity of competition in the model varies from ``permanent monopoly" (just one firm) to ``\gadelete{early entry}\gaedit{incumbent}" (the first-mover in temporary monopoly) to permanent duopoly to ``\gadelete{late entry}\gaedit{entrant" (late-arriver in temporary monopoly)}. \gaedit{We find that, even in the permanent duopoly, in equilibrium competition incentivizes firms to play a ``greedy algorithm" that does not explicitly explore and even more so if the firm is a late arriver in a market.} This algorithm also prevails under monopoly, simply because it tends to be easier to deploy. \gaedit{Whereas the incumbent in the temporary monopoly is incentivized to deploy a more advanced exploration algorithm that perform better in the long run. As a result, we find that consumer welfare is highest under temporary monopoly and that even increasing the number of firms in the market past a duopoly makes consumers weakly worse off.}

@@

\gaedit{The findings on temporary monopoly shed light on the ``first-mover advantage" phenomenon in the digital economy. Being first in the market in the digital economy gives firms free data to learn from (a ``data advantage") as well as a more definite, and possibly better reputation compared to an entrant (a ``reputation advantage").

Even though, under our model, consumers are best off when there is a ``temporary" monopoly, a recent subject of debate in policy circles is whether the data advantage alone poses a sufficiently powerful barrier to entry to warrant anti-trust interest. While \cite{varian2018artificial} and \cite{bajari2018impact} argue that, from a statistical perspective, data has decreasing returns to scale and therefore does not serve as a significant competitive advantage, they focus on the problem in isolation. We argue that, to reason about this question, one needs to consider what the impact of better data is when data collection has competitive consequences, especially when firms need to learn while competing. Therefore, we run additional experiments that attempt to isolate the impact of the data advantage and reputation advantage. We find that the data advantage alone still leads to a significant incumbency advantage. Additionally, the data advantage tends to be stronger when the incumbent commits to a more advanced bandit algorithm. This highlights that, when thinking about data as a barrier to entry, data quality and not just quantity matters. In particular, the data advantage may be larger and more relevant when the incumbent possesses data that it gets via exploration of potentially sub-optimal actions that would be costly for the entrant to gather upon entry.}

@@

The study of whether data can give a significant competitive advantage has been discussed theoretically in \cite{varian2018artificial, lambrecht2015can} and empirically in \cite{bajari2018impact}. These studies consider the effect of additional data in terms of helping towards solving a machine learning problem in isolation and argue that it does not play a large role. We ask how additional data helps while firms simultaneously solve a machine learning problem and compete for consumers. In our setting, the competition dynamics from our model lead to different conclusions about the competitive advantage that a data advantage can give a firm. Additionally, the first-mover advantage has been well-studied in economics and marketing \cite{kerin1992first}. }

### 2018-10-31 paper/content/inverted_u.tex
fix typos
updates to paper

\documentclass[../competing_bandits.tex]{subfiles}
\begin{document}
\section{Competition and Better MAB Algorithms}\label{sec:competition}

@@

Our main experiments are with the duopoly game defined in Section~\ref{sec:model}. As the ``intensity of competition" varies from permanent monopoly to ``incumbent" to permanent duopoly to ``entrant", we find a stylized inverted-U relationship as in Figure~\ref{fig:inverted-U}. More formally, we look for equilibria in the duopoly game, where each firm's choices are limited to \DG, \DEG and \TS. We do this for each ``intensity level" and each MAB instance, and look for findings that are consistent across MAB instances. For cleaner results, we break ties towards less advanced algorithms (as they tend to have lower adoption costs \cite{MWT-WhitePaper-2016,DS-arxiv}). Note that \DG is trivially the dominant strategy under permanent monopoly.

@@

\subsubsection{Inverted-U relationship.}
We interpret our findings through the lens of the inverted-U relationship between the ``intensity of competition" and the ``quality of technology". The lowest level of competition is monopoly, when \DG wins out for the trivial reason of tie-breaking. The highest levels are permanent duopoly and ``entrant" (temporary monopoly from the entrant's perspective). We see that \DG is incentivized for exploration-disadvantaged MAB instances. In fact, incentives for \DG get stronger when the model transitions from permanent duopoly to ``entrant".%
\footnote{For the Heavy-Tail instance, \DG goes from a weakly dominant startegy to a strictly dominant one. For the Uniform instance, \DG goes from a Nash equilibrium strategy to a weakly dominant one.}
Finally, the middle level of competition, ``incumbent" (\gaedit{duopoly with an initial temporary monopoly}) creates strong incentives for \TS. In stylized form, this relationship is captured in Figure~\ref{fig:inverted-U}.

@@

\subsubsection{Death spiral effect.}
Further, we investigate the ``death spiral" effect mentioned in the Introduction. Restated in terms of our model, the effect is that one firm attracts new customers at a lower rate than the other, and falls behind in terms of performance because the other firm has more customers to learn from, and this gets worse over time until (almost) all new customers go to the other firm. With this intuition in mind, we define  \textit{effective end of game (\Eeog)} for a particular \MRV and realization table, as the last round $t$ such that the agents at this and previous round choose different firms. Indeed, the game, effectively, ends after this round. We interpret low \Eeog as a strong evidence of the ``death spiral" effect. Focusing on the permanent duopoly scanario, we specify the \Eeog values in Table \ref{sim_table} (the second line of each cell). We find that the \Eeog values are indeed small:

@@

What do our results mean for the welfare of consumers? In our model we define consumer welfare in terms of pseudo-regret. Following the bandits literature we define pseudo-regret as being the difference between the true mean of the best action and the true mean of the action that is selected. We equate consumer welfare with \textit{market} pseudo-regret where market pseudo-regret is the sum of the pseudo-regret of the actions that are experienced in the competition game\footnote{Note that smaller pseudo-regret is better for consumers}.

@@

Figure \ref{eq_regret} displays the simulated welfare when the equilibrium strategies are played. Consumer welfare is \textit{better off} in the temporary monopoly case than in the duopoly case. However, this was expected from the fact that the incumbent is incentivized to play \TS. The welfare in the temporary monopoly case is close to the welfare to having a single firm playing \TS.

@@

To test if it is possible that increasing competition past a duopoly leads to better welfare than under temporary monopoly we run simulations where we increase the number of firms playing \DG. Figure \ref{many_firm_welfare} reports the average welfare across these simulations. Welfare not only does not get better, \textit{but is even worse} as we increase the number of firms. Correspondingly, we track the average \Eeog in each of the simulations and notice that it is \textit{increasing} as the number of firms increases. As we increase the number of firms, a single individual firm gets fewer consumers for learning as consumers switch between firms more often. Additionally, bad luck with rewards or arm selection is punished even more harshly as we increase the number of firms. This further strengthens the intuition from our previous result, which is that learning may be hard under competition since accidental exploration and mistakes are punished harshly but are necessary for learning.}

>>>>

\documentclass[../competing_bandits_with_appendix.tex]{subfiles}
\begin{document}
\section{Competition and Better MAB Algorithms}\label{sec:competition}

@@

Our main experiments are with the duopoly game defined in Section~\ref{sec:model}. As the ``intensity of competition" varies from permanent monopoly to ``incumbent" to permanent duopoly to ``late entrant", we find a stylized inverted-U relationship as in Figure~\ref{fig:inverted-U}. More formally, we look for equilibria in the duopoly game, where each firm's choices are limited to \DG, \DEG and \TS. We do this for each ``intensity level" and each MAB instance, and look for findings that are consistent across MAB instances. For cleaner results, we break ties towards less advanced algorithms (as they tend to have lower adoption costs \cite{MWT-WhitePaper-2016,DS-arxiv}). Note that \DG is trivially the dominant strategy under permanent monopoly.

@@

\subsubsection{Inverted-U relationship.}
We interpret our findings through the lens of the inverted-U relationship between the ``intensity of competition" and the ``quality of technology". The lowest level of competition is monopoly, when \DG wins out for the trivial reason of tie-breaking. The highest levels are permanent duopoly and ``late entrant". We see that \DG is incentivized for exploration-disadvantaged MAB instances. In fact, incentives for \DG get stronger when the model transitions from permanent duopoly to ``late entrant".%
\footnote{For the Heavy-Tail instance, \DG goes from a weakly dominant startegy to a strictly dominant one. For the Uniform instance, \DG goes from a Nash equilibrium strategy to a weakly dominant one.}
Finally, the middle level of competition, ``incumbent" in the temporary monopoly creates strong incentives for \TS. In stylized form, this relationship is captured in Figure~\ref{fig:inverted-U}.

@@

\subsubsection{Death spiral effect.}
Further, we investigate the ``death spiral" effect mentioned in the Introduction. Restated in terms of our model, the effect is that one firm attracts new customers at a lower rate than the other, and falls behind in terms of performance because the other firm has more customers to learn from, and this gets worse over time until (almost) all new customers go to the other firm. With this intuition in mind, we define  \textit{effective end of game} (\Eeog) for a particular \MRV and realization table, as the last round $t$ such that the agents at this and previous round choose different firms. Indeed, the game, effectively, ends after this round. We interpret low \Eeog as a strong evidence of the ``death spiral" effect. Focusing on the permanent duopoly scenario, we specify the \Eeog values in Table \ref{sim_table} (the second line of each cell). We find that the \Eeog values are indeed small:

@@

What do our results mean for the welfare of consumers? In our model we define consumer welfare in terms of pseudo-regret. Following the bandits literature we define pseudo-regret as being the difference between the true mean of the best action and the true mean of the action that is selected. We equate consumer welfare with \textit{market} pseudo-regret where market pseudo-regret is the sum of the pseudo-regret of the actions that are experienced in the competition game.\footnote{Note that smaller pseudo-regret is better for consumers}

@@

Figure \ref{eq_regret} displays the average welfare when the equilibrium strategies are played. Consumer welfare is \textit{better off} in the temporary monopoly case than in the duopoly case. However, this was expected from the fact that the incumbent is incentivized to play \TS. The welfare in the temporary monopoly case is close to the welfare to having a single firm playing \TS.

@@

To test if it is possible that increasing competition past a duopoly leads to better welfare than under temporary monopoly we run simulations where we increase the number of firms playing \DG. Figure \ref{many_firm_welfare} reports the average welfare across these simulations. Welfare not only does not get better, \textit{but is weakly worse} as we increase the number of firms. Correspondingly, we track the average \Eeog in each of the simulations and notice that it is \textit{increasing} as the number of firms increases. As we increase the number of firms, a single individual firm gets fewer consumers for learning as consumers switch between firms more often. Additionally, bad luck with rewards or arm selection is punished even more harshly as we increase the number of firms. This further strengthens the intuition from our previous result, which is that learning may be hard under competition since  exploration and mistakes are punished harshly but are necessary for learning.}

### 2018-10-31 paper/content/model.tex
fix typos
updates to paper

From each firm's perspective, the interactions with agents follow the protocol of the multi-armed bandit problem (MAB). We focus on i.i.d. Bernoulli rewards: the  reward of each arm $a$ is drawn from $\{0,1\}$ independently with expectation $\mu(a)$. The mean rewards $\mu(a)$ are the same for all rounds and both firms, but but initially unknown.

Before the game starts, each firm commits to an MAB algorithm, and uses this algorithm to choose its actions. Each algorithm receives a ``warm start": additional $T_0$ agents that arrive before the game starts, and interact with the firm as described above. Each firm's objective is to maximize its market share: the fraction of users who chose this firm.

In some of our experiments, one firm is the ``incumbent" who enters the market before the other (``entrant"), and therefore enjoys a \emph{temporary monopoly}. Formally, the incumbent enjoys additional $X$ rounds of the ``warm start". We treat $X$ as an exogenous element of the model, and study the consequences for a fixed $X$.

\subsubsection{Agents.}
Agents are myopic and non-strategic: they choose among the firms so as to maximize their expected reward, without attempting to influence the firms' learning algorithms or rewards of the future users. Agents are not well-informed: they only receive a rough signal about each firm's performance before they choose a firm, and no other information.

Concretely, each firm has a \emph{reputation score}, and each agent's choice is driven by these two numbers. We posit a version of rational behavior: each agent chooses a firm with a maximal reputation score (breaking ties uniformly). The reputation score is simply a sliding window average: an average reward of the last $M$ agents that chose this firm.

@@

For concreteness, we fix one algorithm from each class. Our pilot experiments indicate that the results do not change substantially if other algorithms are chosen. For technical reasons, we consider Bayesian versions initialized with a ``fake" prior (\ie not based on actual knowledge). We consider:

>>>>

From each firm's perspective, the interactions with agents follow the protocol of the multi-armed bandit problem (MAB). We focus on i.i.d. Bernoulli rewards: the  reward of each arm $a$ is drawn from $\{0,1\}$ independently with expectation $\mu(a)$. The mean rewards $\mu(a)$ are the same for all rounds and both firms, but initially unknown.

Before the game starts, each firm commits to an MAB algorithm, and uses this algorithm to choose its actions. Each algorithm receives a ``warm start": additional $T_0$ agents that arrive before the game starts, and interact with the firm as described above. \gaedit{The warm start serves to vary the initial information the firm has before beginning to compete for users as well as to initialize the agents' perception of the firms.} Each firm's objective is to maximize its market share: the fraction of users who chose this firm.

In some of our experiments, one firm is the ``incumbent" who enters the market before the other (``late entrant"), and therefore enjoys a \emph{temporary monopoly}. Formally, the incumbent enjoys additional $X$ rounds of the ``warm start". We treat $X$ as an exogenous element of the model, and study the consequences for a fixed $X$.

\subsubsection{Agents.}
\gaedit{For simplicity we suppose that firms compete for agents on a single dimension, quality, that is summarized via the rewards experienced by past consumers.\footnote{One can think that there are two quality types - a reward of 1 is a high type and a reward of 0 is a low type}} Agents are myopic and non-strategic: they choose among the firms so as to maximize their expected reward \gaedit{(i.e. select the firm with the highest quality)}, without attempting to influence the firms' learning algorithms or rewards of the future users. Agents are not well-informed: they only receive a rough signal about each firm's performance before they choose a firm, and no other information.

Concretely, each firm has a \emph{reputation score}, and each agent's choice is driven by these two numbers. We posit a version of rational behavior: each agent chooses a firm with a maximal reputation score (breaking ties uniformly). The reputation score is simply a sliding window average: an average reward of the last $M$ agents that chose this firm.\gaedit{\footnote{One can think of this as similar to ratings aggregations on many online platforms such as Yelp, Tripadvisor, etc.}}

@@

For concreteness, we fix one algorithm from each class. Our pilot experiments indicate that the results do not change substantially if other algorithms are chosen. \gaedit{We make this restriction since we are mainly interested in qualitative differences between standard ``good" multi-armed bandit algorithms and what algorithms are incentivized in competition.} For technical reasons, we consider Bayesian versions initialized with a ``fake" prior (\ie not based on actual knowledge). We consider:
